Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-12-13 ms.tex
Changing phrasing of "incorrect treatment", etc, to give more specifics about what is wrong

Approximate analytic solutions for resonant scattering exist in certain limits. \citet{1973MNRAS.162...43H} showed that when most of the radiation is out on the damping wings, the transfer equation reduces to the Poisson equation. However, their solution contains an incorrect treatment of the boundary condition. To our knowledge, the errors introduced by this treatment have never been quantified. They attempt a separation of variables as $J(\tau,\sigma) = \theta(\tau) j(\sigma)$ in spatial variable $\tau$ and frequency variable $\sigma$ (their Equations 16 and 23). The solutions for the eigenfunctions $\theta(\tau)$ and $j(\sigma)$ then depend explicitly on the separation constant $\lambda$. In order to satisfy the boundary conditions, the separation constant is shown to satisfy an eigenvalue equation of the form
\be
\lambda \tan(\lambda B) & = & \frac{3}{2} \phi \Delta,
\label{eq:evalue}
\ee
where $2B$ is the slab optical depth at line center, $\phi$ is the line profile, and $\Delta$ is the Doppler width. The key point is that the line profile depends on one of the coordinates: frequency. This causes the eigenvalues of the separation constant to be frequency-dependent. Thus, the separation ``constant'' is not constant, and the function does not satisfy the Poisson equation since the frequency derivatives will act on the separation ``constant", giving extra terms. In the limit of large optical depth $B$, they approximate the eigenvalues as $\lambda_n B \simeq \pi (n-1/2)$. However, this gives zero mean intensity at the surface. Their Equation 34 subsequently allows $\lambda$ to have a small deviation from the above expression, which is explicitly frequency dependent. This allows a nonzero intensity at the surface, but at the cost of rendering the separation of variables assumption invalid. Our treatment, using the correct boundary condition, quantifies the errors in their ansatz.

Several other works have followed \citet{1973MNRAS.162...43H}, building upon their result without addressing the incorrect boundary condition. \citet{1990ApJ...350..216N} attempts to extend the solution to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line, but again does not use the correct boundary condition. \citet{2006ApJ...649...14D} generalize the same problem to spherical geometry, as is used here.

@@

We have examined previous solutions to \lya transfer including resonant scattering in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \citet{1973MNRAS.162...43H}, \citet{1990ApJ...350..216N}, \citet{2006ApJ...649...14D} and others produces a discrepancy in the outgoing spectrum as compared with Monte Carlo. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition. The key result is that the errors in the previously-cited works have been quantified via a correction term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}/H_0$ is of order unity when the spectral peaks are near the Doppler core, and diminishes at larger $\tau_0$ following a $(a\tau_0)^{-1/3}$ scaling.

@@

%TODO: where the last equality (see e.g. \citealt{1994ApJ...427..603R}) is valid on the damping wing.

>>>>

Approximate analytic solutions for resonant scattering exist in certain limits. \citet{1973MNRAS.162...43H} showed that when most of the radiation is out on the damping wings, the transfer equation reduces to the Poisson equation. However, their solution contains an incorrect treatment of the boundary condition. To our knowledge, the errors introduced by this treatment have never been quantified. They attempt a separation of variables as $J(\tau,\sigma) = \theta(\tau) j(\sigma)$ in spatial variable $\tau$ and frequency variable $\sigma$ (their Equations 16 and 23). The solutions for the eigenfunctions $\theta(\tau)$ and $j(\sigma)$ then depend explicitly on the separation constant $\lambda$. In order to satisfy the boundary conditions, the separation constant is shown to satisfy an eigenvalue equation of the form
\be
\lambda \tan(\lambda B) & = & \frac{3}{2} \phi \Delta,
\label{eq:evalue}
\ee
where $2B$ is the slab optical depth at line center, $\phi$ is the line profile, and $\Delta$ is the Doppler width. The key point is that the line profile depends on one of the coordinates: frequency. This causes the eigenvalues of the separation constant to be frequency-dependent. Thus, the separation ``constant'' is not constant, and the function does not satisfy the Poisson equation since the frequency derivatives will act on the separation ``constant", giving extra terms. In the limit of large optical depth $B$, they approximate the eigenvalues as $\lambda_n B \simeq \pi (n-1/2)$, which gives zero mean intensity at the surface. Their Equation 34 subsequently allows $\lambda$ to have a small deviation from the above expression, which is explicitly frequency dependent. This allows a nonzero intensity at the surface, but at the cost of rendering the separation of variables assumption invalid. Our treatment, using the correct boundary condition, quantifies the errors in this ansatz.

Several other works have followed \citet{1973MNRAS.162...43H}. \citet{1990ApJ...350..216N} extends the solution to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line. \citet{2006ApJ...649...14D} generalize the same problem to spherical geometry, as is used here. Both works, however, use the same surface boundary condition and ansatz as \citet{1973MNRAS.162...43H}.

@@

We have examined previous solutions to \lya transfer including resonant scattering in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \citet{1973MNRAS.162...43H}, \citet{1990ApJ...350..216N}, \citet{2006ApJ...649...14D} and others produces a discrepancy in the outgoing spectrum as compared with Monte Carlo. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition. The key result is that the errors in the previously-cited works have been quantified via a correction term, $H_{\rm bc}$, which explains an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}/H_0$ is of order unity when the spectral peaks are near the Doppler core, and diminishes at larger $\tau_0$ following a $(a\tau_0)^{-1/3}$ scaling.

@@
